My first thanks go to my supervisor, Bob Atkey.
The combination of his own knowledge and his understanding of the lacks in my
knowledge has let me have a positive and fulfilling experience as a PhD student.
The wider Mathematically Structured Programming group at Strathclyde has also
taught me much, well beyond what appears in this thesis.
The group has also often been the source of fun and excitement in research,
which I see as crucial to progress.

Special thanks go to Guillaume Allais, Jan de Muijnck-Hughes, and James McKinna,
all of whom have read all or part of this thesis in draft form and whose
comments have lead to great improvements in the final document.
Similar thanks go to the anonymous reviewers of the papers that have lead to and
formed part of this thesis.
More recently, I also thank my examiners Kathrin Stark and Conor McBride,
whose keen and careful reading have lead to yet more improvements to this final
version of my thesis.

The programming language research community as a whole has helped me get to
where I am today.
Interactions at conferences and other meetings have brought me plenty of new
perspectives.
Additionally, starting even before my PhD studies, the formerly Twitter, now
Fediverse, community of researchers and practitioners have been crucial in
getting me interested in programming languages and various more specific topics
thereof.

I thank my family, particularly for our time together in the COVID-19
lockdown/work-from-home period.
The pandemic caused difficulties for us all, but a positive approach got us
through it and let us take advantage of what we could.
More recently, I thank Ayaka, who has listened to and supported me over the last
year of write-up.
Alas, the writing process can take a toll on relationships, so I wish her all
the best.

I have made many friends outside academia during my studies while living in
Glasgow and, latterly, Edinburgh.
Particularly, those of the Strathclyde Board Games Society and Japanese language
learning groups have given me much needed fun times in evenings and weekends, as
well as deeper friendships.
I owe much of my good mood and motivation to carry on to these people.

The work described in this thesis would not have happened without the Agda
programming language/proof assistant and both the contents and style developed
in its standard library.
I thank all contributors to these tools, as well as everyone who has offered
help on how to use them.
Similarly, the preparation of this thesis and related papers was much aided by
\TeX, \LaTeX, and the various packages needed to practically produce beautiful
mathematical documents.

This document is adapted from the template by Jethro Browell
(\url{https://www.overleaf.com/latex/templates/thesis-template-for-university-of-strathclyde/nfnrnmjqyxqg}),
which was licensed under CC BY 4.0.
