\chapter{Conclusions}\label{sec:conc}

In this thesis, I have developed a foundation for semiring-annotated calculi
presented in natural deduction style.
I have given a consolidated account of the semiring-annotated calculus $\name$,
including its relations to existing linear and modal calculi.
As part of this, I have adapted what I call \emph{bunched connectives} from
\cite{RPKV20} as a way to state the typing rules of the calculus as well as to
work with the metatheory.
The distinction between sharing and separating conjunction given by the bunched
connectives corresponds well to the notions of \emph{additive} and
\emph{multiplicative} connectives in the object language, respectively.
Following this, I have given a novel linear algebra-based definition of
\emph{environment} for semiring-annotated calculi, together with a motivation
which may serve as a basis for the corresponding definition for other
substructural systems.
The adequacy of this definition of environment is shown by my implementation of
simultaneous renaming and substitution, as well as other operations on
environments, like composition of renamings and substitutions.
The definition of environments, and hence simultaneous renaming/substitution, is
novel, but acts like the appropriate generalisation of environments for simply
typed theories.

With the details of $\name$ worked out, I then moved on to adapting the work of
\citet{AACMM21} so as to make it able to capture semiring-based usage
restrictions, as found in $\name$.
The syntax descriptions of the resulting system are based on the bunched
connectives, and are shown to be expressive enough to encode calculi of a
variety of forms.
In adapting the semantics, I am forced to be precise about \emph{sharing} and
\emph{separating} bunched connectives, but largely add these as a refinement of
the work of \citet{AACMM21}.
I provide the renaming and substitution operations for all expressible calculi.
I also provide more specialised examples of semantic traversals:
a usage elaborator, an NbE algorithm, a denotational semantics, and translations
between different calculi.
The usage elaborator gives an unexpected example of generic programming, which
one could not write without syntax descriptions.
Meanwhile, the NbE algorithm gives further justification that I have the correct
notion of environment for semiring-annotated type systems.
Together, the examples show the applicability and versatility of the framework
I have developed.

\section{Future work}

To conclude, I discuss various possible future directions of the work started in
this thesis.

\paragraph{Equality}
Perhaps the most fundamental missing piece from the metatheoretic account of
semiring-annotated calculi I have given in this thesis is equations between
terms.
Reasoning about equality between terms and environments is a problem I have
tried to solve, but I have not arrived at a satisfactory solution in the time
available to me.

I believe that the basic difficulty of giving an account for equality in a
linear setting is the proliferation of $\Sigma$-types.
For example, describing equality between two applications of
\TirName{$\with$-I} is immediate: $\Gamma \vdash (M, N) = (M', N') : A \with B$
if and only if $\Gamma \vdash M = M' : A$ and $\Gamma \vdash N = N' : B$.
However, to do the same with \TirName{$\otimes$-I} requires us to be careful
about the contexts of the subterms.
The two applications of \TirName{$\otimes$-I} may a priori split the context in
different ways, and should only be equated when those splittings are equal
(in the appropriate sense).
If the splittings are equal, then the contexts of the subterms will line up, and
only then can the subterms themselves be compared for equality.
These multiple stages come about because $\plr{T * U}\,\Gamma$ is a
$\Sigma$-type, and equality of elements of $\Sigma$-types always follows this
pattern.
Such reasoning becomes even more complex for environments, which are equivalent
to iterated $*$-families.
Additionally, it is unclear what effect subsumption of contexts (like subusaging
in this thesis or explicit structural rules in other calculi) should have on
equality, particularly when the subsumption commutes with parts of the subterms.

Previous work on linear logic has preferred to deal with equations only on
untyped terms. \todo{Check this.}

\paragraph{Polymorphism}
An important feature of most contemporary statically typed programming languages
is polymorphism.
In particular, parametric polymorphism over types can be used to significantly
improve code reuse, and is well undestood theoretically via System $F$ and its
variants.
I have not considered polymorphism in this thesis, and neither did
\citet{AACMM21} in their paper, so whether it can be supported in the framework
presented earlier is an open question.
However, both \citet{Autosubst15} and \citet{Autosubst18} have applied similar
work to polymorphic calculi via special support, suggesting that it would be
possible to modify the framework of this thesis in a similar way.

A separate but related question concerns polymorphism over usage annotations.
The status of polymorphism over usage annotations is less well established both
in practice and in theory.
\Citet{Granule18} present an implementation allowing for polymorphism of usage
annotations, and even polymorphism over semirings, but provide no more than
example programs to justify the feature.
This thesis provides no advance on understanding annotation polymorphism, unless
it can be encoded into a semiring to fit the framework.

%\subsection{Further generalisations}

\paragraph{Structure of contexts}
As I have presented it, the work of \citet{AACMM21} has two axes in which it is
generic: the syntax, which can be controlled through syntax descriptions to
produce a wide range of calculi and features; and the semantics, where we can
produce a wide range of maps out of terms with the help of environments.
To this, the work of this thesis has added a third axis of genericity: the usage
discipline of variables, as described by a partially ordered semiring.

Starting at least with the bunched connectives in \cref{sec:lnd}, if not earlier
when talking about usage contexts forming modules over the semiring of
annotations, I have made productive use of abstractions over the basic usage
annotations throughout this thesis.
These abstractions suggest a next step of completely abstracting away much of
the representation of contexts and their individual entries.
One may imagine that it is possible to develop a framework in which the required
operations and properties of contexts are axiomatised, similar to how usage
annotations are axiomatised to form a semiring in this thesis, and to how
categories-with-families models are defined~\citep{Dybjer95,CCD19}.
Instances of such a framework would include the work of \citet{RPKV20}, which
uses a very similar bunched connective abstraction over a very different
representation of contexts, based on relational interleaving of lists.

The use of semirings is motivated in this thesis and elsewhere largely because
they are general enough to cover a wide range of examples.
However, I cannot claim to have a derivation from first principles of why we
should choose partially ordered semirings over any of a range of similar
algebraic structures.
Additionally, some of the specific constructions done in this thesis fit
somewhat unnaturally with the semiring-based approach.
For example, when translating semiring-annotated systems to traditional systems,
I tended to need to make a \emph{bottom-up} assumption
(\cref{def:DILL-bottom-up,def:PD-bottom-up}) so as to avoid some ``junk'' facts
given by the semiring.
Meanwhile, the usage elaborator of \cref{sec:usage-elaborator} eschews the
``forward'' computation of semiring operations in favour of non-deterministic
backwards computation, e.g., from a sum to the collection of possible summands.
Possibly consciously working more abstractly, as described in the previous
paragraph, would make a more natural structure appear.

If we are to retain an annotation-based approach to usage restrictions, then a
possible feature request that falls out of the encoding of linear/non-linear
logic is to have some sort of kinding system by which different kinds of types
are annotated using different sets of annotations.
In the L/nL example, we would want linear types to be annotated with $\gr0$ and
$\gr1$, and intuitionistic types to be annotated with $\gr\omega$ (as the sole
element of a trivial instance of an algebraic structure), with no crossover
between the two kinds.
Algebraic means to handle such mixed-kind usage vectors may be inspired by the
work of \citet{Hart95,MF21} on dimensional analysis in linear algebra.

\paragraph{Partiality}\label{sec:part}
\input{part.tex}

\paragraph{A separate variable sort}
\Citet{AACMM21} mention as a limitation of their work the fact that variables
can have unrestricted type.
This limitation carries over to the work of this thesis.
In many languages --- for example, the $\mu\tilde\mu$-calculus of
\cref{sec:mu-mu-tilde} and the normal/neutral forms of a $\lambda$-calculus ---
we would like to be able to restrict the syntax so as to disallow variables of
certain kinds.
Such a restriction would make it easier to construct traversals over these
syntaxes, which I have not attempted in this thesis.

In private communication, Allais has suggested a simple solution to this problem
by which the type of variable types and the type of term types are distinct, and
related by an arbitrary relation $R$.
Each use of the \AgdaInductiveConstructor{`var} constructor then requires a
proof that $R$ relates the type of the variable picked to the type of the
conclusion desired.
For example, in the syntax of normal/neutral forms, we let the variable types be
just the types, while the term types are types tagged with whether the term is
normal or neutral.
The relation $R$ relates equal types where the term type is marked neutral.
Allais also suggests that such a scheme could be used to bind patterns in the
context, with the relatedness proof serving as a path through the pattern to a
variable of the desired type.
It remains non-trivial to work out the appropriate notion of renaming, and more
generally the appropriate notion of environment, in this setting, which I leave
to future work.

\paragraph{Logical frameworks}
I discussed in \cref{sec:lin-sys} two previous approaches to representing linear
languages in logical frameworks.
The work of this thesis should help in adapting the first approach --- making a
new logical framework with direct support for linearity --- to a broader range
of substructural calculi.
In particular, the bunched premise connectives give candidates for the required
type formers in such a logical framework.
On the other hand, the second approach --- to use an existing non-linear logical
framework but include a \texttt{linear} predicate --- should be easy to adapt to
semiring-annotated systems by replacing the \texttt{linear} predicate by a
relation between semiring elements and \texttt{term -> term} weak functions.

\paragraph{Fitch-style systems}
I mentioned in \cref{sec:fitch} an alternative way to present modal logics,
namely Fitch-style natural deduction systems.
These appear to be largely irreconcilable with the work of this thesis due to
the fact that the $\Box$-elimination rule is very sensitive to the context.
A general treatment of Fitch-style syntaxes, comparable to the treatment of
semiring-indexed syntaxes given in this thesis, would probably have to assume a
$\Box$ connective whose behaviour was determined as part of the structural rules
of the calculus.
