\chapter{Renaming and substitution for $\name$}\label{sec:ren-sub-lr}

In \cref{sec:semirings}, I defined my calculus of interest $\name$.
In this chapter, I develop the necessary syntactic metatheory for
specifying and implementing the simultaneous substitution operation.
My aim in this chapter is to replay the development in \cref{sec:kits}, making
minimal changes to lemmas and theorems so as to support semiring usage
annotations.
The big change I make in this chapter is to the definition of
\emph{environment}.
In terms of ideas, I believe that the definition of environment I give in
\cref{sec:lrkits} is the main contribution of this thesis.
Its primary novel feature is the use of a linear map --- in the sense found in
linear algebra --- to distribute usage annotations across the values in the
environment.

The resulting substitution operation bears some similarity to a multi-variable
substitution operation given by \citet{petricek-thesis}.
This is the only other presentation I have found in the literature of an
operation similar to simultaneous substitution for a linear or linear-like
calculus.
I discuss Petricek's substitution lemma in more detail in \cref{sec:petricek}.
In brief, relative to Petricek's work, I extract the environment from the
substitution lemma, tidy up the definition, and present usage-annotated
environments as a reusable tool and object of study in their own right.
I reuse the resulting definition of environment in the remainder of this thesis,
taking advantage of the properties and generality of definition established in
this chapter, particularly in \cref{sec:semantics,sec:example-semantics}.

The body of this chapter starts with a motivation for and a definition of
environment in the usage-annotated setting in \cref{sec:lrkits}.
I then give some key properties of and operations on these environments in
\cref{sec:lenv}.
I apply the previous definitions and properties to the problems of simultaneous
renaming and substitution for $\name$ in \cref{sec:lrsub}, which also connects
the previous sections' mathematical developments with Agda code.
I then compare the yielded substitution operation with the one given by Petricek
in \cref{sec:petricek}, before a concluding discussion in
\cref{sec:ren-sub-lr-conc}.

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Comparison with Petricek's substitution lemma}\label{sec:petricek}
\input{petricek.tex}

\section{Conclusion}\label{sec:ren-sub-lr-conc}

In this and the preceding chapter, I have developed a discipline for specifying
the syntax of linear and modal type systems, and furthermore developing the
syntactic metatheory of those type systems.
All of these are based on semirings, and the linear algebra arising from
considering a usage context full of semiring elements as a vector.

These developments can be seen in retrospect as a generalisation of the methods
explained in \cref{sec:simple}.
In terms of premise connectives in the syntactic rules, we have generalised from
just $\{\dot1, \dottimes\}$ to
$\{\dot1, \dottimes, I^*, \sep, \gr r\cdot{}, \Box^{0{+}}\}$, maintaining our
ability to keep the context implicit.
Similarly to how rule premises can require separation of usage annotations, our
new environments can require such a separation between their entries thanks to
the linear map they now contain.
I have generalised the key property of a kit from arbitrary weakening to
weakening by $\gr0$-annotated variables, and using that have produced a
substitution operation based on the same principles as that from
\cref{sec:kits}.

Having generalised all of the components --- namely the contexts, the syntax,
and the notion of environment --- the type of the substitution operation looks
the same as it did for intuitionistic ST$\lambda$C\@.
Being able to maintain this uniformity is a key step towards generalising the
rest of \cref{sec:simple} (i.e., \cref{sec:gen-sem,sec:gen-syn}), as I do in
\cref{sec:framework}.

Future work may want to extend the work of this chapter, in which case
there are some unanswered questions.
Principal among these, in my mind, is dealing with equivalence/equality of
environments.
We want to talk about equality of environments for two related purposes.
The most immediate is that we want to develop the equational theory of renaming
and substitution --- for example, when we use \cref{thm:env-comp} to compose
substitutions, we expect that composition to be associative and unital (with
respect to \cref{thm:env-id}).
These equations of substitutions should yield equations on terms in which such
substitutions have been applied.
Slightly more abstractly, I would like to develop a theory of
\emph{quantitative multicategories}, in which multimorphisms have in their
domain a list of objects paired with usage annotations.
I would hope for $\name$ types and terms to give an example quantitative
multicategory, analogously to how Lambek calculus gives an example of an
ordinary multicategory and the simply typed $\lambda$-calculus gives an example
of a Cartesian multicategory.

Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
$\rho\,x = \sigma\,x$.
This follows from what we expect of equality of functions (function
extensionality).
Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
dividing up the usage annotations of $\Gamma$, and then a function producing
$\V$-values whose usage annotations come from that division.
Equality of $\Sigma$-types is tricky --- we need to equate the first components,
then rewrite the types of the second components by this equation before equating
them.
In practice, the equations rewriting other equations build up so much that
I have given up on a first effort to give a treatment of the equational theory
of substitutions.
Note that recursively defined environments (\cref{def:lr-rec-env}) are also
$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
definition does not clearly help.

I hope that people working on substructural type systems in the future can take
inspiration from the process laid out in \cref{sec:lrkits} when working out the
appropriate notion of environment for their discipline.
Particularly, \cref{def:lr-rec-env} (the recursive definition) should serve as a
specification, if not the actual implementation, when coming across a new
substructural discipline.
As for the progression to \cref{def:lr-env}, this appears to arise from the fact
that the quantitative usage information is a refinement of the intuitionistic De
Bruijn index-based syntax.
%Being just a refinement means that the action of an environment on a variable
%is largely as it was in the intuitionistic case,
